Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
edits for specific comments

%%%%%%%%%%%%
\subsubsection*{Estimates of run-time complexity}

The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
simplified Wright-Fisher model with recombination.

@@

%% Progressive memory reduction as a function of T
To get an idea of how required space depends on the length of time between simplification steps,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

@@

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$,
there would also be around $\mu 2NT$ mutations (and the same number of sites).
Since mutations fall as recombinations do on the marginal trees,
adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
and the same savings as we obtained from edges above,
even without considering the
computational burden of propagating mutations forwards across generations.

>>>>

%%%%%%%%%%%%
\subsubsection*{Computational complexity}

% I may need to cite my student's t-shirt for the following line.
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  
In this section, we explore the how required computational effort and storage space scale
in a Wright-Fisher model with recombination.

@@

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

@@

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$,
there would also be around $\mu 2NT$ mutations (and the same number of sites).
Since mutations fall as recombinations do on the marginal trees,
the number of mutations is proportional to the total length of all the marginal trees,
and so adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
and the same savings as we obtained from edges above,
even without considering the
computational burden of propagating mutations forwards across generations.

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a tree sequence above,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
so $M$ will typically be $O(N \log N)$ as well.
